Blueprint for efficient nuclear spin characterization with color center

### 2024-05-17 PRB - revision reply/main_revision.tex
v2
new
my changes

Where $\vec{n}_0 \cdot \vec{n}_1 = \frac{\omega_L^2+(s_0+s_1)A_{zz}+s_0 s_1 (A_{zz}^2+A_{zx}^2)}{\omega_0 \omega_1}$, \textcolor{blue}{which is the inner product of precession axes of a nuclear spin condition on the electron spin sublevel. This expression indicates the advantage of using DD sequence for sublevels where either $s_0$ or $s_1$ is zero, only first order parallel hyperfine coupling $\frac{A_{zz}}{\omega_L}$ play a role. On the other hand, in spin-1/2 systems ($s_0=-\frac{1}{2}$ and $s_1=\frac{1}{2}$), the first order hyperfine coupling vanishes and only second order parallel hyperfine coupling $(\frac{A_{zz}}{\omega_L})^2$ and second order perpendicular hyperfine coupling $(\frac{A_{zx}}{\omega_L})^2$ remain present. This indicates weak sensitivity of DD sequence on different nuclear spins.}

@@

Although the Dynamical Decoupling (DD) sequence is well understood and used to characterize nuclear spins, it is not applicable to all systems.
First, DD only works for defects in high magnetic fields and low hyperfine coupling ($\omega_L \gg A_{zz}, A_{zx}$).
Second, the signal is limited to the electron spin $T_2$, which only gradually approaches $T_1$.
Third, signal analysis is rather complicated and time-consuming; one has to collect enough data to ensure that no two nuclear spins overlap in the signal.
Fourth, this sequence does not work for spin 1/2 systems, such as group IV defects in diamond, since the parallel component of the hyperfine coupling cancels their effect on the first order electron spin signal.
Hence, the resonance signal depends on the second order of the perpendicular hyperfine component.
\begin{figure*}[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figure_2.pdf}
		\caption{Simulated FFT of ESEEM signal with a zoom into weakly coupled nuclear spins for the register reported \cite{abobeih2019atomic} consist of 23 nuclear spins, assuming the electron spin $S=1/2$ system. b. Enhancing the sensitivity to weakly coupled spins by increasing number of $\pi$-pulses to $N=72$ pulses in each entangling period. $\tau$ is set to the Larmor bright spot. Simulated Fisher Information 56 by 56 matrix for the case of the electron spin c. $S=1$ d. $S=1/2$ system. The first 23 hyperfine parameters are $A_{zz}$ and the second 23 parameters are $A_{zx}$ of the nuclear spin register. \textcolor{blue}{It indicates the covariance between two parameters. The main diagonal appears as each parameter is correlated with itself. The side-diagonals indicate the two hyperfine parameters that belong to the same nuclei.}}
		\label{fig:ddeseem}
	\end{center}
\end{figure*}

In this work, we consider the 23-nuclear spin cluster characterized previously \cite{abobeih2019atomic} as a realistic case for our comparison.
\textcolor{blue}{Each nuclear spin is identified with two hyperfine parameters $A_{zz}$ and $A_{zx}$, which shows coupling to the central electron spin. Hence, characterizing this 23-nuclear spins requires identifying 56 hyperfine coupling parameters.}
First, we apply the Dynamical Decoupling (DD) sequence to this nuclear spin register and visualize the obtained results. Fig \ref{fig:3} compares the DD signal for the same nuclear spin register but different defects in diamond.
An intuitive description of the sensitivity of a sequence to the variation of hyperfine coupling could be estimated by considering the width and sensitivity of the position of the resonances.
First, we consider the case of an NV center in diamond. The width of a dip is $\delta \tau = \frac{A_{zx}}{\omega_L^2}$. Taking the derivative of the resonance time, we define the sensitivity of the longitudinal hyperfine coupling, using Eq. \ref{eq:resonance} to simplify:
\begin{gather}
	\delta A_{zz} \equiv \frac{\delta \tau}{\abs{\frac{\partial \tau_p}{\partial A_{zz}}}} = \frac{2}{\tau_p} \frac{A_{zx}}{\omega_L} \ge \frac{4}{T_2^{HE}} \frac{A_{zx}}{\omega_L}
\end{gather}
If we assume typical values of $T_2^{HE} = 100$ $\mu$s and $\omega_L=500$ kHz, a weakly coupled nuclear spin $A_{zx}=5$ kHz can be distinguished from another weakly coupled nuclear spin with $\delta A_{zz} = 400$ Hz.
For $S=1/2$, such as group IV defects in diamond, the resonance times are sensitive to the second order of $A_{zx}$ and show no sensitivity to $A_{zz}$.
Hence, even at large $\tau$, the resonances related to nuclear spins will not be well separated.
We define the sensitivity for the spin one-half system as follows:
\begin{gather}
	\delta A_{zx} \equiv \frac{\delta \tau}{\abs{\frac{\partial \tau_k}{\partial A_{zx}}}} = \frac{4}{\tau_k} \ge \frac{8}{T_2^{HE}}
\end{gather}
Assuming typical values of $T_2^{HE} = 100$ $\mu$s, a nuclear spin can be distinguished from another one if their transverse hyperfine coupling is separated by $\delta A_{zx} = 80$ kHz, which is quite inaccurate and approximately 160 times worse than for $S=1$.
\begin{figure*}%[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figure_3.png}
		\caption{a. FFT of ESEEM signal for $\tau$ from 10 $\mu$s to 500 $\mu$s with a step of 10 $\mu$s b. 2D correlation of each frequency for different $\tau$.}
		\label{fig:4}
	\end{center}
\end{figure*}

>>>>

Where $\vec{n}_0 \cdot \vec{n}_1 = \frac{\omega_L^2+(s_0+s_1)A_{zz}+s_0 s_1 (A_{zz}^2+A_{zx}^2)}{\omega_0 \omega_1}$, \textcolor{blue}{which is the inner product of precession axes of a nuclear spin condition on the electron spin sublevel. This expression indicates the advantage of using DD sequence for sublevels where either $s_0$ or $s_1$ is zero, only first order parallel hyperfine coupling $\frac{A_{zz}}{\omega_L}$ play a role. On the other hand, in spin-1/2 systems ($s_0=-\frac{1}{2}$ and $s_1=\frac{1}{2}$), the first order perturbation correction with respect to hyperfine coupling vanishes and only second order perturbation term with respect to parallel $(\frac{A_{zz}}{\omega_L})^2$  and perpendicular coupling $(\frac{A_{zx}}{\omega_L})^2$ remain present. This indicates weak sensitivity of DD sequence on different nuclear spins}.

@@

Although the Dynamical Decoupling (DD) sequence is well understood and used to characterize nuclear spins, it is not applicable to all systems.
First, DD only works for defects in high magnetic fields and low hyperfine coupling ($\omega_L \gg A_{zz}, A_{zx}$).
Second, the signal is limited to the electron spin $T_2$, which only gradually approaches $T_1$.
Third, signal analysis is rather complicated and time-consuming; one has to collect enough data to ensure that no two nuclear spins overlap in the signal.
Fourth, this sequence does not work for spin 1/2 systems, such as group IV defects in diamond, since the parallel component of the hyperfine coupling cancels their effect on the first order electron spin signal.
Hence, the resonance signal depends on the second order of the perpendicular hyperfine component.
\begin{figure*}[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figure_2.pdf}
		\caption{Simulated FFT of ESEEM signal with a zoom into weakly coupled nuclear spins for the register reported \cite{abobeih2019atomic} consist of 23 nuclear spins, assuming the electron spin $S=1/2$ system. b. Enhancing the sensitivity to weakly coupled spins by increasing number of $\pi$-pulses to $N=72$ pulses in each entangling period. $\tau$ is set to the Larmor bright spot. Simulated  \textcolor{blue}{5pESEEM} Fisher Information 56 by 56 matrix for the case of the electron spin c. $S=1$ d. $S=1/2$ system. The first 23 hyperfine parameters are $A_{zz}$ and the second 23 parameters are $A_{zx}$ of the nuclear spin register. \textcolor{blue}{It indicates the covariance between two parameters. The main diagonal appears as each parameter is correlated with itself. The side-diagonals indicate the two hyperfine parameters that belong to the same nuclei.}}
		\label{fig:ddeseem}
	\end{center}
\end{figure*}

In this work, we consider the 23-nuclear spin cluster characterized previously \cite{abobeih2019atomic} as a realistic case for our comparison.
\textcolor{blue}{Each nuclear spin is identified with two hyperfine parameters $A_{zz}$ and $A_{zx}$, which shows coupling to the central electron spin. Hence, characterizing this 23-nuclear spins requires identifying 56 hyperfine coupling parameters.}
First, we apply the Dynamical Decoupling (DD) sequence to this nuclear spin register and visualize the obtained results. Fig \ref{fig:3} compares the DD signal for the same nuclear spin register but different defects in diamond.
An intuitive description of the sensitivity of a sequence to the variation of hyperfine coupling could be estimated by considering the width and sensitivity of the position of the resonances.
First, we consider the case of an NV center in diamond. The width of a dip is $\delta \tau = \frac{A_{zx}}{\omega_L^2}$. Taking the derivative of the resonance time, we define the sensitivity of the longitudinal hyperfine coupling, using Eq. \ref{eq:resonance} to simplify:
\begin{gather}
	\delta A_{zz} \equiv \frac{\delta \tau}{\abs{\frac{\partial \tau_p}{\partial A_{zz}}}} = \frac{2}{\tau_p} \frac{A_{zx}}{\omega_L} \ge \frac{4}{T_2^{HE}} \frac{A_{zx}}{\omega_L}
\end{gather}
If we assume typical values of $T_2^{HE} = 100$ $\mu$s and $\omega_L=500$ kHz, a weakly coupled nuclear spin $A_{zx}=5$ kHz can be distinguished from another weakly coupled nuclear spin with $\delta A_{zz} = 400$ Hz.
For $S=1/2$, such as group IV defects in diamond, the resonance times are sensitive to the second order of $A_{zx}$ and show no sensitivity to $A_{zz}$.
Hence, even at large $\tau$, the resonances related to nuclear spins will not be well separated.
We define the sensitivity for the spin one-half system as follows:
\begin{gather}
	\delta A_{zx} \equiv \frac{\delta \tau}{\abs{\frac{\partial \tau_k}{\partial A_{zx}}}} = \frac{4}{\tau_k} \ge \frac{8}{T_2^{HE}}
\end{gather}
Assuming typical values of $T_2^{HE} = 100$ $\mu$s, a nuclear spin can be distinguished from another one if their transverse hyperfine coupling is separated by $\delta A_{zx} = 80$ kHz, which is quite inaccurate and approximately 160 times worse than for $S=1$.
\begin{figure*}%[htbp]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{figure_3.pdf}
		\caption{a. FFT of ESEEM signal for $\tau$ from 10 $\mu$s to 500 $\mu$s with a step of 10 $\mu$s b. 2D correlation of each frequency for different $\tau$.}
		\label{fig:4}
	\end{center}
\end{figure*}

### 2024-05-17 PRB - revision reply/reply.tex
my changes

	9) Could the authors provide a short comment comparing to entanglement
	based approaches for nuclear spin control, e.g.,
	https://www.nature.com/articles/ncomms14660?

@@

	\color{black} \section*{Reply to the First Referee}\\
	We appreciate the comments made by the referee. We clarify each point as follows:\\

	1. Second order hyperfine couplings means terms like $(\frac{A_{zz}}{\omega_L})^2$ and $(\frac{A_{zx}}{\omega_L})^2$. This is stated in the sentence:
	((... the second-order term $\frac{A_{zx}^2}{\omega_L^2}$ can be neglected ... ))

	To clarify this point, we add the following explanation to the text:\\
	((... \textcolor{blue}{Each nuclear spin is identified with two hyperfine parameters $A_{zz}$ and $A_{zx}$, which shows coupling to the central electron spin. Hence, characterizing this 23-nuclear spins requires identifying 56 hyperfine coupling parameters.} ...)) \\

	((... \textcolor{blue}{which is the inner product of precession axes of a nuclear spin condition on the electron spin sublevel. This expression indicates the advantage of using DD sequence for sublevels where either $s_0$ or $s_1$ is zero, only first order parallel hyperfine coupling $\frac{A_{zz}}{\omega_L}$ play a role. On the other hand, in spin-1/2 systems ($s_0=-\frac{1}{2}$ and $s_1=\frac{1}{2}$), the first order hyperfine coupling vanishes and only second order parallel hyperfine coupling $(\frac{A_{zz}}{\omega_L})^2$ and second order perpendicular hyperfine coupling $(\frac{A_{zx}}{\omega_L})^2$ remain present. This indicates weak sensitivity of DD sequence on different nuclear spins}...))\\

@@

	4. The Hamiltonian in Eq. 1 is very well-understood in this community [ref here]. The Hamiltonian conventionally is written in the rotating frame of applied microwave and consist of Electron and nuclear spin Zeeman energy and dipole interaction between them. As it is mentioned in the text, the secular approximation is considered so terms like $S_x, S_y$ do not appear. Traditionally, the nuclear spins' frame are rotated such that $I_y$ term do not appear. This Hamiltonian applied to any two-level system, e.g., this could be any two spin sublevels in Silicon Carbide (S=3/2)\\

	5. We thank the referee for pointing out labels for subfigures (pdf is corrected but latex does not show). We fixed this issue. In Fig. 3, we are discussing 5pESEEM sequence mostly for Electron spin 1/2 system. We did not include FFT of 5pESEEM sequence for spin 1 system because it works just as good as DD sequence, and does not bring any additional information. However, the Fisher information obtained for spin 1 system in Fig. 3c is good metrics to compare with the result of electron spin 1/2 system in Fig 3d.  \\

	6. The initial state for the Ramsey sequence can be any of the two sublevels of spin system. Indeed, the gray exponential decay is not sufficient to describe spin relaxations as mentioned in the text, but it shows the time scale of the decay. We thank the referee for this point and we decided to remove this exponential decay to avoid confusion. \\

@@

	8. This equations are valid as long as the condition stated before Eq. 1 are valid. Most importantly secular approximation is crucial, and also the nuclear spin should be spin 1/2 like Carbons. The Spin of electron spin is not play an important role since they can be well separated and pulses resonant with only an effective two level system can be considered.\\

	9. In this work, we address nuclear spins only with microwave driving of the electron spin (did we mention this in the text?). Applying another radio frequency (RF) driving adds another experimental challenge to the problem. In this case, RF spectroscopy to find and drive individual nuclear spins is required, which can add the required time to image nuclear spins.\\

	\color{red} \section{ Report of the Second Referee}
	The authors propose a blueprint for efficient nuclear spin
	characterization with color centers. The authors claim that they
	propose a more straightforward approach for determining the hyperfine
	interactions among each nuclear and the electron spin. The manuscript
	is written clearly by comparing dynamical decoupling (DD) and Electron
	Spin Echo Envelop Modulation (ESEEM), the numerical results are valid
	and may could be of use in other future experiments. I think the
	results obtained by the authors are of interest for Phys. Rev. B
	readers, so I support the publication in its present form.

@@

	2. Indeed, the gray exponential decay is not sufficient to describe spin relaxations as mentioned in the text, but it shows the time scale of the decay. We thank the referee for this point and we decided to remove this exponential decay to avoid confusion.\\

	3.

@@

	\color{black} \section*{ Reply to the Third Referee}
	We appreciate the comments made by the referee. We clarify each point as follows:\\
	1. Characterization of spin-1 system with both DD and DDESEEM works similarly well. Indeed, it is important to observe the diagonal and side-diagonals are distinguishable from in the Fisher Information matrix, which in both Fig. 2c and 3c are clear. However, the color scale is levelled such that in each sequence, the perfomance of spin 1/2 can be compared with spin 1. Fig. 3c, 3d are Fisher Information matrix simulated for 5pESEEM. For clarity we added this point in the caption.\\

>>>>

	9) Could the authors provide a short comment comparing to entanglement
	based approaches for nuclear spin control, e.g. : \\
	https://www.nature.com/articles/ncomms14660?

@@

	\color{black} \section*{Reply to the First Referee}
	We appreciate the comments made by the referee. We clarify each point as follows:\\

	1. We thank the Referee for the usefull comment, indeed we used a confusing  formulation. Second order hyperfine couplings mean terms like $(\frac{A_{zz}}{\omega_L})^2$ and $(\frac{A_{zx}}{\omega_L})^2$ in the resonance conditions. This is stated in the sentence:
	((... the second-order term $\frac{A_{zx}^2}{\omega_L^2}$ can be neglected ... ))
	We replaced this terms by "second order correction of the resonance with respect to hyperfine parameters", which in our opinion better reflects the meaning. 
	To clarify this point, we add the following explanation to the text:\\
	((... \textcolor{blue}{Each nuclear spin is identified with two hyperfine parameters $A_{zz}$ and $A_{zx}$, which shows coupling to the central electron spin. Hence, characterizing this 23-nuclear spins requires identifying 56 hyperfine coupling parameters.} ...)) \\

	((...\textcolor{blue}{which is the inner product of precession axes of a nuclear spin condition on the electron spin sublevel. This expression indicates the advantage of using DD sequence for sublevels where either $s_0$ or $s_1$ is zero, only first order parallel hyperfine coupling $\frac{A_{zz}}{\omega_L}$ play a role. On the other hand, in spin-1/2 systems ($s_0=-\frac{1}{2}$ and $s_1=\frac{1}{2}$), the first order perturbation correction with respect to hyperfine coupling vanishes and only second order perturbation term with respect to parallel $(\frac{A_{zz}}{\omega_L})^2$  and perpendicular coupling $(\frac{A_{zx}}{\omega_L})^2$ remain present. This indicates weak sensitivity of DD sequence on different nuclear spins}...))\\

@@

	4. \textcolor{red}{We need to either add a reference or a small paragrah saying why Secular is applied and a $U_{int}$ to rotation frame, now it is too rude answer}The Hamiltonian in Eq. 1 is very well-understood in this community [ref here]. The Hamiltonian conventionally is written in the rotating frame of applied microwave and consist of Electron and nuclear spin Zeeman energy and dipole interaction between them. As it is mentioned in the text, the secular approximation is considered so terms like $S_x, S_y$ do not appear. Traditionally, the nuclear spins' frame are rotated such that $I_y$ term do not appear. This Hamiltonian applied to any two-level system, e.g., this could be any two spin sublevels in Silicon Carbide (S=3/2)\\

	5. We thank the referee for pointing out labels for subfigures (pdf is corrected but latex does not show). In the revised version of the manuscript we fixed this issue. 
	In Fig. 3, we are discussing 5pESEEM sequence mostly for Electron spin 1/2 system. 
	We did not include FFT of 5pESEEM sequence for spin 1 system because it works just as good as DD sequence, and does not bring any additional information. 
	However, we believe, that the Fisher information obtained for spin 1 system in Fig. 3c is good metrics to compare with the result of electron spin 1/2 system in Fig 3d.  \\

	6. \textcolor{red}{Initial state should be initialised into superposition of two eigenstates, s0,s1, which can be any, right?}The initial state for the Ramsey sequence can be any of the two sub-levels of spin system. Indeed, the grey exponential decay is not sufficient to describe spin relaxations as mentioned in the text, but it shows the time scale of the decay. We thank the referee for this point and we decided to remove this exponential decay to avoid confusion. We have added this correction to the main text.
	\textcolor{blue}{((..initial state is XXX..))}.\\

@@

	8. This equations are valid as long as the condition stated before Eq. 1 are valid. Most importantly secular approximation is crucial, and also the nuclear spin should be spin 1/2 like Carbons. The spin of electron spin is not playing an important role since various allowed microwave transitions can be well separated and pulses resonant with only an effective two level system can be considered.\\

	9. In this work, we address nuclear spins only with microwave driving of the electron spin (\textcolor{red}{did we mention this in the text?}). Applying another radio frequency (RF) driving adds another experimental challenge to the problem and goes beyond the scope of this work. In this case, RF spectroscopy to find and drive individual nuclear spins is required, which can add the required time to image nuclear spins, which was successfully demonstrated in various works for spin 1, 3/2,  \cite{DDRF, ENDOR, SSRV2}. We believe that it will as well work well for spin $S=1/2$, as it probes the transition frequency of the nuclear spin in $m_s=1/2$ or $m_s=-1/2$ subdomains, making it sensitive to the first order of the $A_{zz}$ coupling. Comparison of various RF assisted protocols goes beyond the scope of the current work and requires further investigation.\\

	\color{red} \section{Report of the Second Referee}
	The authors propose a blueprint for efficient nuclear spin
	characterization with color centers. The authors claim that they
	propose a more straightforward approach for determining the hyperfine
	interactions among each nuclear and the electron spin. The manuscript
	is written clearly by comparing dynamical decoupling (DD) and Electron
	Spin Echo Envelop Modulation (ESEEM), the numerical results are valid
	and may could be of use in other future experiments. I think the
	results obtained by the authors are of interest for Phys. Rev. B
	readers, so I support the publication in its present form.

@@

	2. Indeed, the grey exponential decay is not sufficient to describe spin relaxations as mentioned in the text, but it shows the time scale of the decay. We thank the referee for this point and we decided to remove this exponential decay to avoid confusion.\\

	3. We thank the referee for noting this and fixed the references in the revised version of the manuscript.

@@

	\color{black} \section*{ Reply to the Third Referee}
	We appreciate the comments made by the referee. We clarify each point as follows:\\

	1. Characterization of spin-1 system with both DD and DDESEEM works similarly well. Indeed, it is important to observe the diagonal and side-diagonals are distinguishable from in the Fisher Information matrix, which in both Fig. 2c and 3c are clear. However, the color scale is levelled such that in each sequence, the perfomance of spin 1/2 can be compared with spin 1. Fig. 3c, 3d are Fisher Information matrix simulated for 5pESEEM. For clarity we added this point in the caption.\\
